---
title: "Comparing lightgbm to other R GBDT libraries"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
---

```{r setOpts, echo = FALSE}
knitr::opts_chunk$set(
  eval = FALSE
  , echo = TRUE
  , warning = FALSE
  , message = FALSE
  , tidy = FALSE
)
```

# Intro to `{lightgbm}`

The examples in this demo use the `nycflights13::weather` dataset, to solve the following regression problem:

> Given observations of other weather signals, what is the expected humidity?

## Setting Up Training Data

### Creating a Dataset

```{r createDataset}
library(lightgbm)
library(nycflights13)

# convert from data frame to matrix
feature_names <- c(
    "month"
    , "day"
    , "hour"
    , "temp"
    , "dewp"
    , "wind_dir"
    , "wind_speed"
    , "precip"
    , "pressure"
)
target_name <- "humid"

X <- as.matrix(weather[, feature_names])
y <- weather[[target_name]]

dtrain <- lightgbm::lgb.Dataset(
    data = X
    , label = y
    , colnames = feature_names
)
```

### Customizing Dataset Construction

The parameter `categorical_feature` can be used to tell LightGBM which features to treat as categorical. This should result in a better-performing model than would be produced if LightGBM treated those columns as continuous.

```{r categoricalData}
library(data.table)
library(lightgbm)
library(nycflights13)

weatherDT <- data.table::as.data.table(nycflights13::weather)

# convert from data frame to matrix
feature_names <- c(
    "month"
    , "day"
    , "hour"
    , "temp"
    , "dewp"
    , "wind_dir"
    , "wind_speed"
    , "precip"
    , "pressure"
)
target_name <- "humid"

X_train <- as.matrix(weatherDT[month <= 11, .SD, .SDcols = feature_names])
y_train <- weatherDT[month <= 11,][[target_name]]

# hold out December data
X_test <- as.matrix(weatherDT[month == 12, .SD, .SDcols = feature_names])
y_test <- weatherDT[month <= 12,][[target_name]]

dtrain <- lightgbm::lgb.Dataset(
    data = X_train
    , label = y_train
    , colnames = feature_names
    , categorical_feature = c("month", "day", "hour")
)
```

### Creating a new Dataset from a "reference" Dataset

Once a `Dataset` object has been set up, you can create other `Dataset` objects from new raw data but using the same bins and other preprocessing rules from construction of that initial dataset.

This is especially useful for the creation of evaluation datasets.

TODO

## Training

### Choosing an Objective Function

See [the LightGBM parameter documentation](https://lightgbm.readthedocs.io/en/latest/Parameters.html#objective) for a complete list of built-in regression objectives (sometimes referred to as "loss functions").

The examples below use mean absolute error (L1 loss).

```{r maeLoss}
library(data.table)
library(lightgbm)
library(nycflights13)

weatherDT <- data.table::as.data.table(nycflights13::weather)

# convert from data frame to matrix
feature_names <- c(
    "month"
    , "day"
    , "hour"
    , "temp"
    , "dewp"
    , "wind_dir"
    , "wind_speed"
    , "precip"
    , "pressure"
)
target_name <- "humid"

X_train <- as.matrix(weatherDT[month <= 11, .SD, .SDcols = feature_names])
y_train <- weatherDT[month <= 11,][[target_name]]

# hold out December data
X_test <- as.matrix(weatherDT[month == 12, .SD, .SDcols = feature_names])
y_test <- weatherDT[month == 12,][[target_name]]

dtrain <- lightgbm::lgb.Dataset(
    data = X_train
    , label = y_train
    , colnames = feature_names
    , categorical_feature = c("month", "day", "hour")
)

mod <- lightgbm::lgb.train(
  data = dtrain
  , obj = "regression_l1"
  , nrounds = 100L
)
```

### Using a Custom Objective

### Categorical Features

### Customizing Tree Growth

### Adding Validation Sets

### Using Multiple Evaluation Metrics

### Using Early Stopping

### Using Custom Metrics

## Inspecting Models

## Tuning Hyperparameters

### Model Structure

### Feature Importance

## Deployment

# Other Libraries

## `{catboost}` examples

## `{xgboost}` examples
